Seismic parameter estimation and the Canadian crust

### 2013-09-12 paper/draftBP.tex
near end of discussion A

The Slave, Superior and Churchill provinces share similar mean crustal compositions, with the Slave province being slightly more tonalitic in average composition. These three provinces cluster around the average for the Canadian Shield. The Grenville Province on the other hand differs significantly in mean composition. One interpretation is that the crust in the Grenville province has a mean crustal composition corresponding to a more mafic lower crustal composition than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the South-East of the province also have seismic ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). The average $V_P/V_S$ value for the Grenville Province is 1.77 marginally higher than the average value computed by the study of Eaton et. al. [2005].

We now take the

[ Continue with crustal composition discussion ]

@@

Whereas Zandt and Ammon show increasing $V_P/V_S$ values with crustal age the estimates computed in this study indicate an opposite trend, that is lower seismic velocity ratios with crustal age. This aligns more closely to the results of Durrheim and Mooney [1991] who provide results showing a thick higher velocity Proterozoic crust compared to a thinner lower velocity Archean crust. The crustal thickness data attained in this study also show thicker crusts for Proterozoic units (38.9 km) than Archean units (38.0 km) though this difference is not as drastic as the greater than 5km difference reported by Durrheim and Mooney. Though $Vp$ data was not explicity estimated in this study, the higher seismic velocity ratio in the Proterozoic units (1.76) than in Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. Though Durrheim and Mooney accept the hypothesis that Proterozoic crust is more prone to basaltic underplating, thereby explaining higher velocities and thicker crust, Roberta Rudnick [1995; Rucknick and Gao, 2003] has argued persuasively that a range of possible processes may explain this data.

[ Use crustal composition plus increasing Vp/Vs with younger crust result to make ``best'' pick among the 4 main processes Rudnick proposes to complete section.]

>>>>

The estimates for Canada and the Canadian Shield are, to a first-order approximation, located between two candidate end-members, granite geniss and diorite. This intermediate or andesitic composition is in general agreement with other recent studies on crustal composition [Rudnick and Fountain, 1995; Christensen and Mooney [1995]. The Slave, Superior and Churchill provinces share similar mean crustal compositions, with the Slave province being slightly more tonalitic in average composition. These three provinces cluster around the average for the Canadian Shield. The Grenville Province on the other hand differs significantly in mean composition. One interpretation is that the crust in the Grenville province has a mean crustal composition corresponding to a more mafic lower crustal composition than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the South-East of the province also have seismic ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). The average $V_P/V_S$ value for the Grenville Province is 1.77 marginally higher than the average value computed by the study of Eaton et. al. [2005].

\begin{table}
  \begin{tabular}{ p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} }
    \hline
    Oxide, wt\% & Smithson [1978] & Rudnick and Fountain [1995] & Christensen and Mooney [1995] & Rudnick and Gao [2003] & This Work \\
    \hline
    $SiO_2$   & 63.0 & 52.6 & 61.7 & 60.6 & 64.3 \\
    $TiO_2$   & 0.7 & 0.8 & 0.9 & 0.7 & 0.6 \\
    $Al_2O_3$ & 15.8 & 16.6 & 14.7 & 15.9 & 15.5 \\
    $Fe_2O_3$ & 2.0 & --- & 1.9 & --- & 1.9 \\
    $FeO$     & 3.4 & 6.6 & 5.1 & 6.7 & 3.3 \\
    $MnO$     & 0.1 & 0.1 & 0.1 & 0.1 & 0.1 \\
    $MgO$     & 2.8 & 4.4 & 3.1 & 4.7 & 2.2 \\
    $CaO$     & 4.6 & 6.4 & 5.7 & 6.4 & 4.3 \\
    $Na_2O$   & 4.0 & 3.2 & 3.6 & 3.1 & 3.6 \\
    $K_2O$    & 2.7 & 1.9 & 2.1 & 1.8 & 2.9 \\
    $H_2O$    & --- & --- & 0.8 & --- & 1.1 \\
    $P_2O_5$  & --- & 0.2 & 0.2 & 0.1 & 0.2 \\
    $CO_2$    & --- & --- & --- & --- & 0.1 \\
    \hline
  \end{tabular}
  \caption{Bulk crustal composition comparison from selected sources. Rudnick and Fountain [1995] and Rudnick and Gao [2003] use a combined $Fe$ quantity and is listed under $FeO$}
\label{table:composition}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{composition}
  \caption{Comparison of major crustal elements from different studies. All estimates have been normalized to the results computed in this study.}
  \label{fig:composition}
\end{figure}

The $V_P$ and $V_S$ estimate for Canada lies on a line connecting diorite with granite gneiss (\ref{fig:lith}) and this can be used to estimate the bulk crustal composition. Using this analysis the bulk crustal lithology of the Canadian crust is 51\% diorite and 49\% granitic. The elemental breakdown is provided in table \ref{table:composition} along with previously published global crustal studies. The elemental crustal estimates produced in this study show a crust with marginally higher $Sio_2$ composition and lower $Fe$ and $Mg$ content. It is noteworthy that both studies involving greater reliance on seismic analysis, this study and the P-wave velocity study by Smithson [1978] closely follow this trend. Figure \ref{fig:composition} provides the major crustal elements from the data in table \ref{table:composition} but normalizes the estimates to the Candian crustal composition model produced here. Those estimates that are within within 30\% of model are highlighted by the gray shaded region. Also included in this figure is the crustal composition estimate for the Canadian Shield. The estimate for this Archean shield is much closer to the global crustal estimates of Christensen {\it et al.} and Smithson and indeed is bracketed by estimates from these two studies. These data suggest that the extensive Platform and Orogenic regions of Canada skew the bulk composition away from global averages and are comparatively better represented by granite gneiss lithologies than diorites. In contrast, the Canadian Shield crust shares a composition in greater alignment to the global averages which is best represented as 68\% dioritic and 32\% granitic in the formulation given above. Interestingly, the root-mean-squared error between the model proposed here and the other models presented is minimized by choosing a diorite component of 69\%, which results in an rms of 1.19\%.

[[[ Can I actually say any of that? Isn't the data on elemental composition likely one of the main differentiators? The foundations seem a little weak, I don't want to start building castles, thoughts? ]]]

@@

Whereas Zandt and Ammon show increasing $V_P/V_S$ values with crustal age the estimates computed in this study indicate an opposite trend, that is lower seismic velocity ratios with crustal age. This aligns more closely to the results of Durrheim and Mooney [1991] who provide results showing a thick higher velocity Proterozoic crust compared to a thinner lower velocity Archean crust. The crustal thickness data attained in this study also show thicker crusts for Proterozoic units (38.9 km) than Archean units (38.0 km) though this difference is not as evident as the greater than 5km difference reported by Durrheim and Mooney. Though $Vp$ data was not explicity estimated in this study, the higher seismic velocity ratio in the Proterozoic units (1.76) than in Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. The hypothesis that Proterozoic crust is prone to basaltic underplating, thereby explaining higher velocities and thicker crust, suggested by Durrheim and Mooney [1991], has been persuasively augmented by recent work [1995; Rucknick and Gao, 2003] to now include a range of possible processes that may explain the data.

One possible process is the preferential recycling of soluble ions such as $Mg^{2+}$ and $Ca^{2+}$. These ions are weathered and carried into the oceans leaving behind silicon and aluminum [Rudnick and Gao, 2003] producing a net change to crustal composition over time. In another model, deep crustal roots expose lower crust to processes leading to eclogitization and subsequent delamination due to a density inversion with the underlying mantle peridoties. This model does not directly address the question as to why there is an apparent variance in average velocities between younger and older crusts, unless the delamination process has evolved over time. However, knoweldge of the spatial or temporal extent of delamination is difficult to document [Rudnick, 1995]. A third model suggests that a mixture of silicic melts from subducting oceanic crust and basaltic melts from mantle peridotites were important in early crustal formation. This process is thought to be more common in the hotter Archean and is related to the TGG suites of Archean granitoids [Rudnick and Gao, 2003] and therefore provides a possible explanation for the data. It is important to state that all of these processes, as well as others, may have been operating throughout geologic time.

[It is very possible that all of these as well as other processes have been operating thoughout geologic history, [Rudnick, 1995]]

This study provides results that are in general agreement with other recent crustal studies and similarily argues for an intermediate or andesitic crust. There are several processes that have been suggested to explain this intermediate composition ...

... Taking a break ...
